# Using R

## Installing Packages {#packages}

Sometimes you will want to use tools that are not built into the baseR code. You can download these tools from R repositories as `packages`. A package is a collection of functions, data, and documentation designed to accomplish a particular task. An R package can help with tasks such as data visualization, statistical modeling, or data cleaning. You can install an R package through the RStudio interface, or through entering code into the console.

```{r error=FALSE, eval=FALSE}
install.packages("amt") # install a new package
library(amt) # call a package you previously downloaded
```

***

**`EXERCISE`** 

Load Tidyverse package using RStudio

1. Select `Packages`
2. Select `Install`
3. Type `tidyverse`
4. Select `Install`
5. Bonus: Install tidyverse in the code editor using the example code above.

**We will go over what tidyverse is later on in this course**

***

If two different packages contain functions with the same name, R will mask one with the other based on the order they are loaded. To avoid confusion, you can explicitly tell R which package to use with the `package::function()` format. For example, the `filter()` function exists in both `stats` and `dplyr` packages. If you specifically want the version from `dplyr`, you can write `dplyr::filter()` in your code. We will talk more about this later on.

## Using Comments

Use the `#` to tell R you are making a comment. Comments are used to explain code and allow someone unfamiliar with your code to follow more easily. Commenting can also be used to prevent R from running specific lines of code since R ignores anything that follows the `#` mark.

```{r}
# 567*5 tells R that 567*5 is a comment, and so R knows not to execute this line of code.
```

Sometimes you want to comment out large sections of code, and this can be done using `control + shift + c` on windows or `command + shift + c` on a Macbook.

***

**`EXERCISE`**

1. Create a new script (**<em>File > New File > R Script</em>**)
2. In the first line, type `# Intro R`
3. On line 2, type ` 1 + 1`
4. Select `Run` in the upper right corner of the console. You can also use the shortcut `control + enter` on Windows and `command + enter` on a Macbook. If the keyboard shortcuts aren’t working, navigate to **<em>Tools > Modify Keyboard Shortcuts </em>**, type **<em>Run Current Line or Section</em>** into the search bar, double-click the shortcut, remove the current key combination, and enter your preferred shortcut.

***

## Functions {#functions}

A function is a defined script that is used to accomplish a particular task. Functions use an input to give a desired output. R provides built-in functions, such as `mean()`, `sum()`, and `plot()`, which perform commonly used operations. Every function has arguments that determine what kind of inputs are needed to make the function run. 

* Arguments: information that goes inside the parenthesis to tell the function what to do. For example, when we used the `seq` function, the arguments are `from`, `to`, and `by`. Many arguments have a default value, which is a value that is automatically used for an argument if you don’t provide your own. For example, in `seq()`, the default value for the `from` argument is 1.
* Pass: We pass a value to a function argument. We can pass the value 1 to the argument `from`, and the value 10 to the argument `to` and the value 0.1 to the argument `by`.
* Return: This is the terminology to say that the function gives us an output. So with the `seq(from = 1, to = 10, by = 0.1)`, the function <em>returns</em> a sequence of numbers

```{r eval=FALSE}
function_name <- function(arg1, arg2) { # arg1 and arg2 are inputs
  # Code that processes the inputs
  result <- arg1 + arg2
  return(result)  # return() provides the output of the function
}
```

The order of arguments in a function call matters, especially when you don't name them. To avoid confusion, it's often helpful to explicitly name the arguments. You can use the help function `?` to check the correct usage of a function. For example, to learn more about the `rep()` function, you can execute the command `?rep` in your Console.

Additionally, you can define your own functions to solve specific problems and use them repeatedly in your code without having to repeat the code multiple times. For example, if you need to convert elevation from meters to feet in several places in your script, you can write a function to do it once and then use that function whenever you need it. Functions are an essential part of R programming, as they allow for cleaner, more efficient, and modular code. Although we won’t be covering custom functions in this course, it’s important to note that creating your own functions is a common and useful practice in programming.

``` {r eval=FALSE}
# Write a function to convert a column of elevation in meters to feet
convert_elevation_to_feet <- function(elev_m) { # vector of elevation in meters is input (argument)
  elev_ft <- elev_m * 3.28084 # code to process elevation in meters to feet
  return(elev_ft) # return output of function
}
```

***

**`EXERCISE`**

1. Use the `?` function to look up the seq() function. 
2. Why does `seq(from = 1, by = 3)` return a sequence successfully, while `seq(from = 40, by = 3)` results in an error? What is causing the difference in behavior? Hint: look up `seq` using the help function. Are there any default values? How might that affect the behavior of the function?

***

## Use R as a calculator

Remember, order of operations matters. The order is the same as you learned back in school.
  
From highest to lowest precedence:
  
| Operator     | Symbol(s)   | Description                 |
|:--------------|-------------|----------------------------|
| Parentheses  | `(` `)`     | Controls the order of operations |
| Exponents    | `^`, `**`   | Raises a number to a power       |
| Divide       | `/`         | Division                        |
| Multiply     | `*`         | Multiplication                  |
| Add          | `+`         | Addition                        |
| Subtract     | `-`         | Subtraction                     |`

```{r eval=TRUE, include=TRUE}
1 + 100 # performs addition
3 + 5 * 2 # performs multiplication before addition
(3 + 5) * 2 # performs addition in () before multiplication 
2/10000 # division
2 ^ 3 # exponentiation: 2 raised to the power of 3
```

## Use R to compare things
To compare things in R, we use logical operators. Below is a brief list. 

| Operator | Meaning                    |
|:---------|:---------------------------|
| `==`     | is equal to                |
| `>`      | greater than               |
| `<`      | less than                  |
| `>=`     | greater than or equal to   |
| `<=`     | less than or equal to      |
| `!`      | not                        |
| `|`      | or                         |
| `%in%`   | is contained in            |

Let's go through a few examples of using logical operators. Notice how R evaluates each of these lines of code as TRUE or FALSE. We are essentially asking R if the above comparison is TRUE or FALSE. We will go over `%in%` later in this class. 

```{r}
1 == 9  # equality (note two equals signs, read as "is equal to")
1 != 1  # inequality (read as "is not equal to")
1 < 2  # less than
1 <= 1  # less than or equal to
```

***

**`EXERCISE`**

1. In your code editor, enter the following lines of code. Then run them to view each line of code's output in your console. 
    a. 67 * 9
    b. 7 + 9 + 10
    c. 4399 - 871 * (9 + 1) 
2. Evaluate whether the following logical conditions are TRUE or FALSE
    a. `(380*3) == (190*6)`
    b. `567 > 890`
    c. `30 >= (27 + 1 + 2)`
2. Bonus: What is wrong with the following code? 
    a. Type `87(9 + 1)` into your console
    b. Execute that specific line of code by placing the cursor on that line and selecting `Run`. 
    c. What error do you see? What do you think is happening? How might you fix this?  

*** 

## Use R to assign objects {#r-objects}

Objects are a bit of an abstract concept. All you really need to know for now is that objects are things that we make in R that can take on a variety of structures with different data types, and when we assign them a name, they get saved in our global environment. **Objects are data structures with associated data attributes.** 

Object assignment lets us assign a name to a value, making it easier to use later. This helps avoid repeating code. We assign a value to a variable using the assignment arrow `<-` or the `=` so that R recognizes the name as a reference to the object. So when we run, `x <- 6`, it reads <em>make x contain 6 </em> It's recommended to use the `<-` since the `=` can get mixed up with assigning values to function arguments. Once we assign an object to a variable, it is stored in our global environment (upper right hand panel of RStudio).

```{r}
x <- 1/40 # here we are telling R to assign 1/40 to the variable x so that it recognizes x as an object in our global environment
x <-  24 # variables can easily be re-assigned/over-written
y <- x * 2 # We can use existing objeects in expressions to create new objects

rm(y) # you can also remove objects from the environment using the rm() function
```

### Variable names

Variable names can contain letters, numbers, underscores and periods. 
**They CANNOT start with a number OR contain any spaces.** Recall that R is case sensitive.

A few different conventions for longer variable names:

- periods.between.words
- underscores_between_words
- camelCaseToSeparateWords

Your choice of convention is up to you, *JUST BE CONSISTENT*.

***

**`EXERCISE`**

1. Assign `1+56` to a variable called `x1_a`
2. Assign `sqrt(24)` to a variable called `x1_b`. Use `?sqrt` to learn more about the `sqrt()` function. What is the name of the argument of the `sqrt()` function?
3. Bonus: Chained assignments. What happens when you execute `x3.c <- y3.c <- 9/10` into your console? Are `x3.c` and `y3.c` different or equal values? 

***

## Data Types (Modes) {#data-types}

There are 6 main classes of common data modes (i.e. data types): `numeric`, `character`, `logical`, `integer`, `complex`, and `factor`. Data modes refers to the basic type of data stored in an object. For example, numeric mode stores numbers (integers or decimals), and character mode stores text. We typically only use  `factor`, `numeric`, `character`, and `logical` data modes. 

| Mode        | Description                                | Example                    |
|:------------|:-------------------------------------------|:---------------------------|
| factor      | Categorical data (special class of integer with labels stored as characters)| Levels: "Sheep", "Lion", "Deer" (internally stored as: 1, 2, 3)    |
| numeric     | Numbers (includes integers and decimals)   | 3.14, 42                   |
| character   | Text strings                               | "Lion180"                  |
| logical     | TRUE or FALSE values                       | TRUE, FALSE                |

<br>

To ask R what `class` a data mode or object is, we use the `class()` function.

```{r eval=FALSE}
class(1.11) # numeric: any real number
class(1L) # integer: any integer. The L suffix forces the number to be an integer
class(TRUE) # logical: binary TRUE or FALSE 
# You can have data that look essentially the same, but have different classes. 
class('1') # character: words; "" denote words
class(1) # numeric; any real number
class(factor("1a")) # factor: denotes categorical variables, they can be words or numbers
```

You can coerce to a desired data type, as long as they follow the rules
using the functions `as.<desired data type>`

**Coercion Hierarchy (from general to specific): Character → Numeric → Logical**

1. character most general: anything can be turned into a character by adding "quotes" 
    * e.g. as.character(TRUE) → "TRUE"
2. numeric: can read integer and logical types as numbers. Cannot reliably coerce character strings unless they represent valid numbers
    * e.g. "3.14" → 3.14, TRUE → 1; FALSE → 0; "S438" → NA
3. logical: most specific, cannot turn turn character or numeric into a logical type without correctly specifying the value. Coercing a character or invalid numeric value to logical produces an NA
    * e.g. as.logical("sheep" → NA); as.logical(3.14 → TRUE) 
4. Factor: (special case) internally stored as integers with labels. To get the original numeric value, you must coerce factor to character first, then to numeric. Direct coercion from factor to numeric returns internal integer codes (not labels)


```{r}
# create numeric object
a <- 45.6
class(a)

# Convert from numeric to character
a_character <- as.character(a)
class(a_character)

# Create factored object
lion <- factor(c("194", "180"), levels = c("194", "180")) # The levels must match the exact format of the values
lion # print value into console
as.character(lion) # coerced to character
as.numeric(lion) # coerced to numeric
```

You can also coerce from factor to character to numeric in one call

```{r}
as.numeric(as.character(lion))
```

A common mistake in R is using data of the wrong type. In the example below, an error occurs because R cannot convert a character value into a numeric one. It's important to ensure that the data you're working with is of the correct type. 

```{r}
# convert from character to numeric
L262 <- "Lion262"
L262_numeric <- as.numeric(L262)
L262_numeric
```

***

**`EXERCISE`**

1. Assign `"S437"` to a variable called `sheep` (`sheep <- "S437"`). Make sure to include the `""`.
2. What type of data class is sheep? This can be figured out by doing `class(sheep)`
3. Create an object with a value of `45`. Coerce to a character class using `as.character()`.
4. Create the following objects: `S1 <- as.numeric(as.character(1))` and `S2 <- as.logical(as.numeric("1"))`. 
5. Do the values change? Why or why not? Think about this answer in the context of the coercian hierarchy.
6. Bonus: What happens when you execute `sheep <- S437` without `""`? Explain the error. 

***

## Other Object Types

R includes many other object types that we won't cover in detail during this course. For instance, spatial object types are used to represent and work with geographic data. Spatial objects can take on different data types depending on if they are from the `sp` or `sf` package. 

## Data Structure Classes {#data-classes}

Remember when we talked about objects as data with attributes? Well, R offers several ways to store data, depending on what kind of structure you need. The most common types are vectors, data frames, and lists. Each of these can store different types of information and are useful in different contexts when working with data. Think of them as different types of containers for data that are designed to hold and organize data in specific ways. We will be discussing `Scalar`, `Vector`, `Data Frames`, `Matrices`, and `Lists`. Below is a summary of each of these data structures, and we will also go through each of these individually. 

<br>

| Data Structure | Description                                              | Can Contain Different Types? | Dimensions      | Example                  |
|:---------------|:----------------------------------------------------------|:------------------------------|:----------------|:--------------------------|
| Scalar         | A single value (a vector of length 1)                    | No                            | 1 (length = 1)  | `x <- 42`               |
| Vector         | A sequence of elements of the same type                  | No                           | 1 (length > 1)  | `c(1, 2, 3)`             |
| Data Frame     | Table-like structure with columns of equal length        | Yes (by column)              | 2 (rows × cols) | `data.frame(a = 1:3, b = c("x", "y", "z"))` |
| Matrix         | 2D array with all elements of the same type              | No                           | 2 (rows × cols) | `matrix(1:6, nrow = 2)`  |
| List           | A collection of elements that can be of different types  | Yes                          | 1 (named or not)| `list(name = "A", age = 25, scores = c(1,2,3))` |

<br>

### Scalar
A scalar object contains only one element (length is 1). An element is an individual value or item within a data structure

```{r}
x <- 3
```

***

**`EXERCISE`**

1. Create a scalar object named my_scalar and assign it the value 87

***

### Vector
A scalar object is a special case of a vector with a length of 1. A vector in R is essentially a collection of elements **of the same basic data type**. Each 'thing' in the vector is called an element. You can combine values into one vector with different values using the concatenate function `c()`. 

```{r}
my.deer <- c("GDL256", "RVD1011", "CDB567")
```

You can also create an empty vector. If you don’t specify a data type, R will default to logical, or, you can declare an empty vector of whatever type you like.

```{r}
my_vector <- vector(length = 3, mode = "numeric")
my_vector  # this is a logical vector
num_vector <- c(101, 222, 323, 435, 556) # numeric vector
```

What happens when we add elements of different data types to a vector?

```{r}
combine_vector <- c(211, "CDB678", TRUE)
combine_vector
class(combine_vector) 
```

R coerces all the elements in "combine_vector" to character. It can't make `"CDB678"` into a number but it can turn `211` and `TRUE` into text strings (think about the coercion hierarchy)

You can also assign `NA` values to a vector of defined length as well. R is able to handle missing values, and these missing values are given `NA`. When you read in an csv file with empty cells, R will assign these values as NA. A `0` is not the same as `NA`, since R treats `0` as a numeric data class.   

```{r}
# create a vector of 10 NA values
x <- rep(NA, 10)
# set the first element in x to be the number 0
x[1] <- 0
# test if zero is NA
is.na(x)
# R returns a logical vector for TRUE if the value is NA and 
# FALSE for non-NA values
```

We can also make a series of numbers using `:` or the `seq()` function.

```{r}
my_series <- 1:10 # a vector of integers of length 10
my_series
# make series of numbers from 1 to 10 by increments of 0.1
my_seq <- seq(from = 1, to = 10, by = 0.1) 
my_seq
```

***

**`EXERCISE`**

1. Make a character vector of lion ids into a vector called `my_lion`
2. Make a vector of deer ages into a vector called `deer_ages`
3. Create a vector with variable name of your choice with a sequence of numbers from 1 to 100 using both the `seq()` and `:` method
4. BONUS 1: Explain the logic behind coercian that occurs in making this vector. `a.new.vector <- c(FALSE, TRUE, 0)`. Check the class of `a.new.vector` using the `class()` function.
5. BONUS 2: Create the following object: `animals <- factor(c("sheep", "lion", "deer", "deer"), levels = c("sheep", "lion", "deer"))`. Use the `class()` and `str()` functions to examine the `animals` vector. Use the `?` function to understand what `str()` does. What values are assigned to "sheep", "lion", and "deer"? Now convert (coerce) this vector of factors to show each level's label value instead of the internal integer code.

***

### Lists
A list in R is essentially an object with data that can be in different data types/modes. Many functions in R return outputs as lists. 

```{r}
# list with numeric, character, and logical classes
my_list <- list(1, "banana", TRUE)
your_list <- list(2, "apple", FALSE)
my_list
```

We can also append to a list with c() like we did with a vector. However, c() will now treat each element in the vector (here "Lion171" & "S222") as separate list elements to append. 

```{r}
my_new_list <- c(my_list, c("Lion171", "S222")) # here we overwrite the original 
# "my_list" list to include two additional list elements of "Lion171" and "S222".
my_new_list
```

If we want to add one additional list element that contain Lion171 and S222, we have to use the `list()` function in front of concatenate. 

```{r}
my_new_list2 <- c(my_list, list(c("Lion171", "S222"))) # here we overwrite the original 
```

Notice that my_new_list contains 6 elements (Lion171 and S222 are put in different lists) and my_new_list contains 5 elements (Lion171 and S222 are put into the same list).

Sometimes we want to retrieve the values from a certain element in a list. We do this with the `$` operator. The `$` operator is used for indexing named elements in a list. It allows you to access part of a data object for extracting or subsetting data. 

```{r}
my_lions <- list("L211", c("L194", "S789"), "L180")
names(my_lions) # are names automatically assigned? Nope, so let's assign names with the names() function.
names(my_lions) <- c("Lion211", "LionsAndSheep", "Lion180")
my_lions$LionsAndSheep
```

If you want to index a specific element in the list, you can also use brackets.

```{r}
# index second list [[2]] and second element [2] of the second list
my_lions[[2]][2]
```

***

**`EXERCISE`**

1. Create a list of length 3 with at least two different data types. 
2. Use the `class` and `str` functions to explore the contents of your list. 
4. Print the list you create in your Script into the Console. 

***

### Matrix
A matrix in R is a two-dimensional data structure that stores elements of the same type (usually numbers). You can think of it like a spreadsheet or a table with rows and columns, but every value must be the same type (e.g., all numeric or all character). Matrices are structured in terms of rows and columns. For example, a 3x4 matrix has 3 rows and 4 columns.

Matrices are useful when you want to perform mathematical operations on tables of numbers, like multiplying rows and columns, or applying functions across rows or columns. 

```{r}
MyMatrix <- matrix(1:6, nrow = 2, ncol = 3)
MyMatrix
```

***

**`EXERCISE`**

1. Create a 3x1 matrix representing deer population estimates for Round Valley. Assign it the a variable name of your choice.
Each row should represent the total number of individuals in the fawn, yearling, and adult age classes. The column represents a single year. 
2. What does the dim function do? Use `?dim`. Notice dim does not work on vectors. 
3. Use the dim function to print the dimensions of your matrix into the console. `dim(<your matrix object>)`

***

### Data Frames
A data frame in R is a like a list or generalized matrix but with the constraints that:

(1) all list elements are vectors (i.e. they have 1 mode), 
(2) all vectors have the same length
(3) all columns (the list elements) have names

Essentially, imagine each column in the dataframe as a vector, and the dataframe is just a big list of all those vectors. Unlike actual lists however, in dataframes all of the columns/vectors MUST be the same length and have names. Data frames are also also structured in terms of rows and columns.

``` {r}
# Create vectors for each column
sheep_id <- c("S601", "S602", "S603")
sex <- c("F", "M", "F")
age <- c(4, 6, 3) 
capt_date <- as.Date(c("2024-03-15", "2024-03-16", "2024-03-18"))
capt_loc <- c("Acrodectes Peak", "Sawmill Canyon", "Goodale Canyon")
# Combine into a data frame
capt_dat <- data.frame(
  ID = sheep_id,
  Sex = sex,
  Age = age,
  CaptureDate = capt_date,
  CaptureLoc = capt_loc
)

# View the data frame
print(capt_dat)
```

We can use the `$` operator on data frames, just like we do with lists, to extract a specific column by name. For example, `capt_dat$CaptureLoc` will return all the values in the `CaptureLoc` column and print them in the Console.

```{r}
capt_dat$CaptureLoc
```

***

**`EXERCISE`**

1. Create a data frame with three columns and three rows to represent deer mortalities. 
The columns should be named "deer_id", "COM", and "Age". Assign (`<-`) it a variable name of your choice. 
* "deer_id" and "COM" should be character
* "Age" should be numeric
2. What are the dimensions of this data frame? What function can you use to check this?
3. Use the `nrow()` and `ncol()` functions to display the number of rows and columns in your data frame
4. Print the `Sex` column using the `$` operator. Recall that the syntax is `<df>$<column_name>`

***

## Indexing {#index}

Indexing in R refers to the process of accessing specific elements from data structures like vectors, lists, or data frames. The type of indexing you use depends on the structure of your data and the task at hand, making it a versatile tool for data manipulation. To extract elements from a vector, use their index inside square brackets `[]`. 

| **Indexing Method**       | **Description**                                                                                   | **Example**                             |
|---------------------------|---------------------------------------------------------------------------------------------------|-----------------------------------------|
| **Positional Indexing**    | Use numeric positions (1-based) to access elements in a vector, list, or data frame.              | `my_vector[2]` returns the 2nd element. |
| **Named Indexing**         | Access elements by their name, making the code more readable.                                      | `my_vector["b"]` returns the value associated with `b`. |
| **Logical Indexing**       | Use a logical vector (TRUE/FALSE) to subset elements. Only elements corresponding to TRUE are selected. | `my_vector[c(TRUE, FALSE, TRUE)]` returns elements where TRUE is present. |
| **Range Indexing**         | Extract a range or sequence of elements using a colon or vector of indices.                       | `my_vector[2:4]` returns elements from positions 2 to 4. |
| **Negative Indexing**      | Exclude elements by using negative indices.                                                       | `my_vector[-2]` excludes the 2nd element. |
| **which() Indexing**| Find the indices of elements that satisfy a condition, and use those indices for subsetting.      | `which(my_vector > 20)` returns indices where the values are greater than 20. |

### Numerical Indexing

This method involves using the position of elements in a vector, list, or other data structures. R uses 1-based indexing, meaning the first element has an index of 1, not 0 (as in some other programming languages like Python or C).

```{r}
lion_list <- c("194", "228", "267", "272")
lion_list[1] # extract first element out of the lion_list vector
lion_list[4] # extract 4th element out of the lion_list vector
lion_list[c(2:4)] # extract elements 2 to 4 from the my_seq vector
```

### Named Indexing

You can also extract elements by name rather than by index. Using names for indexing is particularly useful when you want to avoid relying on numeric positions, which may change as new data is added.

```{r}
horn_length <- c(10, 30.4, 90, 50.8, 30) # create vector of horn lengths
names(horn_length) <- c("S4", "S567", "S489", "S488", "S89") # assign sheep names to vector
# we can also name a vector 'on the fly'
horn_length <- c(S4 = 10, S567 = 30.4, S489 = 90, S488 = 50.8, S89 = 30) 
horn_length[c("S489", "S488")]
```

### Logical Indexing

Since comparison operators (e.g. >, <, ==) evaluate to logical vectors, we can also use them to succinctly subset vectors: the following statement gives the same result as the previous one. This involves using logical vectors (TRUE/FALSE) to index elements. The logical vector must be of the same length as the data structure you’re indexing. Only elements corresponding to TRUE are selected.

```{r}
horn_length[horn_length > 30] # this statement first evaluates horn_length > 7, generating a logical vector c(FALSE, FALSE, TRUE, FALSE, TRUE), and then selects the elements of horn_length corresponding to the TRUE values.
```

### Range or Sequence Indexing

You can extract a range or sequence of elements using a colon : to specify a range or a vector of indices.

```{r}
horn_length[1:3]
```

### Negative Indexing

Negative indexing is used to exclude elements from a data structure. When you use a negative index, R excludes the element at that specific position. Note that negative indexing only applies to numeric positions, not named elements.

```{r}
horn_length[-2]
# we can skip multiple elements
horn_length[c(-1, -2)]  
# or 
horn_length[-c(1,5)]
# a common mistake would be to ask R x[-1:3] # but there isn't a negative first row
# But remember the order of operations. 
# : is really a function. It takes its first argument as -1, 
# and its second as 3, so generates the sequence of 
# numbers: c(-1, 0, 1, 2, 3).
horn_length[-(1:3)]
# To remove elements from a vector, we need to assign the result 
# back into the variable:
horn_length <- horn_length[-4]
horn_length
```

### Which Indexing

The which() function returns the indices of elements that satisfy a given condition (useful in combination with logical operations).

```{r}
to_remove <- which(horn_length > 30)
to_remove # index of elements to remove from the to_remove vector
horn_length[to_remove]
```

***

**`EXERCISE`**

1. Create the following object. new_deer <- c("RVD1", "CDB222", "EWR343", "WWR526")
2. Print the first three elements of new_deer into the Console using the sequence indexing
3. Remove the second element ("CDB222") using negative indexing, and update the new_deer object.
4. Use the `which()` function to identify the index of the element equal to "WWR526" and print the result.
5. Create the following object. `age <- c(43, 44, 33, 22)`
6. Use logical indexing to print the elements of `age` that are less than 44. Display the result in the Console.

***

### Indexing different data types

Different object types in R require distinct indexing syntax, as each type has its own structure and behavior. For instance, vectors are indexed using [], while lists require [[]] to extract individual elements. Data frames can be subsetted using both [] and [[]], but the syntax varies depending on whether you're accessing rows, columns, or individual elements. Understanding the proper indexing method for each object type is crucial for effectively manipulating and extracting data in R.

| **Data Structure**   | **Indexing with `[]`**                                                                                  | **Indexing with `[[]]`**                                                                                             | **Description**                                                                                          |
|----------------------|---------------------------------------------------------------------------------------------------------|-----------------------------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------|
| **Vector**           | `my_vector[2]` - Accesses the 2nd element.                                                               | Not applicable (use `[]` for vectors).                                                                                  | `[]` returns a subset (vector), and `[]` is used to extract individual elements by their position.       |
| **Data Frame**       | `my_df[2, ]` - Accesses the 2nd row (returns a data frame).                                               | `my_df[[2]]` - Accesses the 2nd column (returns a vector).                                                            | `[]` keeps the result as a data frame, while `[[]]` extracts a specific column or element.               |
| **Matrix**           | `my_matrix[2, 3]` - Accesses the element in the 2nd row and 3rd column.                                 | Not applicable (use `[]` for matrices).                                                                                | `[]` is used to index both rows and columns, but `[[]]` is not used with matrices.                        |
| **List**             | `my_list[2]` - Returns the 2nd element of the list (as a list).                                           | `my_list[[2]]` - Returns the 2nd element of the list (as the actual object, not as a list).                           | `[]` returns a sublist, while `[[]]` extracts the actual object at that index.                           |
| **List of Lists**    | `my_list_of_lists[2]` - Returns the 2nd list in the list of lists (as a list).                           | `my_list_of_lists[[2]]` - Returns the 2nd list (the entire list, not just an element).                               | `[]` returns a sublist, while `[[]]` extracts the entire second list when using a list of lists.         |
| **Nested List**      | `my_list[[2]][[2]]` - Returns the 2nd element of the 2nd list in the list of lists (after first subsetting). | `my_list[[2]][[2]]` - Extracts the 2nd element from the 2nd list in the list.                                         | `[]` can be used to subset lists, and `[[]]` extracts elements within the nested list structure.         |

***

**`EXERCISE`**

1. Create the following object.

* `elk_df <- data.frame(`  
    `elk_id = c("E001", "E002", "E003", "E004", "E005"),`  
    `weight = c(350, 410, 290, 500, 375),`  
    `herd = c("North", "South", "East", "West", "North"))`
    
2. Print rows 1 to 3 and the second column of `elk_df` to the Console. Recall that to index data frames, you use df[rows, columns].
2. Remove the elk ids "E001" and "E002" from the data frame using negative indexing.
3. Use which() indexing to subset the `elk_df` to include only elk from the "North" herd. Assign this to a new object called "elk_north. 
4. BONUS: Use a which() statement to subset `elk_df` to elk in the "North" and "East" herds

***

## BaseR Plotting

BaseR plotting is an efficient way to visualize a dataset. 

```{r}
# Create a fake data frame for bighorn sheep
bighorn_df <- data.frame(
  ID = paste0("S", 500:530),      # ID for each sheep
  Age = sample(1:15, 31, replace = TRUE),  # Random ages between 1 and 10
  Weight = round(rnorm(31, mean = 75, sd = 15), 1),  # Random weights with mean of 75 kg and sd of 15
  Sex = factor(sample(c("M", "F"), 31, replace = TRUE), labels = c("M", "F")),     # Random sexes (Male, Female)
  Herd = sample(c("North", "South", "East", "West"), 31, replace = TRUE)  # Random herd names
)

# Define grey tones based on levels of the 'Sex' factor
grey_tones <- c("blue", "orange")  # grey50 for males, grey80 for females
```


```{r}
# create a scatterplot
plot(x = bighorn_df$Age, # x-axis
     y = bighorn_df$Weight, # y-axis
     main = "Bighorn Sheep: Weight by Age", # title
     xlab = "Age (years)", # x-axis label
     ylab = "Weight (kg)", # y-axis label
     col = grey_tones[bighorn_df$Sex],  # point color (sex needs to be a factor)
     pch = 19,              # solid circle
     cex = 1.2)             # slightly larger points
# Add separate trend lines for each Sex group
# Males
abline(lm(Weight ~ Age, data = subset(bighorn_df, Sex == "M")), col = "blue", lwd = 2)  
# Females
abline(lm(Weight ~ Age, data = subset(bighorn_df, Sex == "F")), col = "orange", lwd = 2) 

# Add a legend
legend("topright", 
       legend = levels(bighorn_df$Sex),  # Use the levels of the factor for legend
       col = c("blue", "orange"),      # Different tones for trend lines
       pch = 16) # solid circle point type
```

```{r}
hist(bighorn_df$Weight,
     main = "Distribution of Bighorn Sheep Weights",
     xlab = "Weight (kg)",
     ylab = "Frequency",
     col = "skyblue",        # fill color of the bars
     border = "black",       # outline color of the bars
     breaks = 10)            # number of bins (adjust as needed)
```

***

**`EXERCISE`**

1. Create the following object: 
```{r eval=FALSE}
# Create a fake data frame for bighorn sheep
bighorn_df <- data.frame(
  ID = paste0("S", 500:530),      # ID for each sheep
  Age = sample(1:15, 31, replace = TRUE),  # Random ages between 1 and 10
  Weight = round(rnorm(31, mean = 75, sd = 15), 1),  # Random weights with mean of 75 kg and sd of 15
  Sex = sample(c("M", "F"), 31, replace = TRUE),     # Random sexes (Male, Female)
  Herd = sample(c("North", "South", "East", "West"), 31, replace = TRUE)  # Random herd names
)
```

2. Create a histogram to visualize the age distribution of bighorn sheep.
* Set the bins to be maroon.
* Add a black border to the bins.
* Include a title and labels for the x and y axes.
* Adjust the breaks (number of bins) as needed

***
