# Data Organization {#data-org}

Let’s start by downloading some data and exploring it in a data frame. In this example, we’ll load a CSV file. 

```{r}
mort <- read.csv("./docs/data/BighornMortTable.csv")
```

## Exploring Data Frames

Let’s take a look at the data. Exploring your dataset is one of the most important first steps when working in R. It helps you understand the structure, types of variables, and potential issues before doing any analysis. A good place to start is by examining the structure of the data to see what you’re working with.

```{r}
# str() shows us the structure of the data, including the data mode, 
# the dimensions of the dataframe including the data mode, and a few observations
str(mort)  
length(mort) # gives the number of columns
nrow(mort) # to get the number of rows
ncol(mort) # number of columns
dim(mort) # or both at once, dim()
names(mort) # names of the columns
```

It's a good time to ask ourselves if the structure of our data matches our intuition:

* Does the data type for each column make sense? 
* If not, we need to sort out those issues now so we don't run into issues down the road.

We can use the functions `head()` and `tail()` to also look at the first or last 6 rows of the data frame. We can adjust the number of rows that are printed using the `n` argument.

```{r}
head(mort, n = 3) # n = refers to how many rows to print 
```

We can also examine individual columns.

```{r}
class(mort$AnimalYear) 
class(mort$Animal_ID)
str(mort$Cause_Mort)
```

We can index a data frame column using the `$` operator and `[`. Below, the `$` operator accesses the `Animal_ID` columns and `[1]` grabs the first value in that column. 

```{r}
mort$Animal_ID[1]
```

Similarly, `[[` will act to extract a single column. 
```{r}
mort[["AnimalYear"]][500] # extract the 500th row of the AnimalYear column
```

We can also use square brackets to extract specific elements from a data frame. Remember that data frame indexing follows the format [rows, columns], so the first number refers to the row, and the second to the column.

```{r}
# [rows, columns]
mort[1,2] # grab element at first row and second column
mort[1,1:6] # grab row 1 in columns 1-6
mort[1:6,5] # grab row 1-6 in column 5
mort[1:6,1:6] # grab row 1-6 in columns 1-6
head(mort[,1:3], n = 5) # grab all rows from columns 1-3, show me the first 5 rows
```

## Vectorized Operations {#vectorized}
R is a natively vectorized language, which means it can automatically perform operations on entire vectors (columns of data) without the need for explicit loops. This makes data manipulation in R efficient and concise. 

For example, let’s say all the ages in our bighorn mortality data set are off by one year. To correct this, we can simply add 1 to the entire Age column. This works because R applies the operation to each element of the vector. But first, we need to make sure the `Age` column is numeric. If it’s stored as character or factor, we need to coerce it to numeric before performing the math. Because R vectorizes math operations, the `+ 1` is automatically applied to every value in the `Age` column, and the result is stored in the new `AgeNew` column — no loop required.

```{r}
class(mort$Age)
mort$AgeNew <- as.numeric(mort$Age)
mort$AgeNew <- mort$AgeNew + 1

```

We see a warning about NAs being introduced when coercing values to a numeric data type. This happens because some entries in the Age column aren’t purely numeric—for example, values like "4+" or ">7" contain extra characters (+ and >). Since R can’t interpret these as numbers, it replaces them with `NA`.

In practice, we could clean this column so it contains only numeric values. One way to do that is by using the `gsub()` function along with regular expression to remove non-numeric characters. I’ll show you the code below for reference, but we won’t be covering regular expressions in detail during this workshop.

```{r}
mort$AgeNew <- gsub("[^0-9.]", "", mort$Age)  # Remove all non-numeric characters
mort$AgeNew <- as.numeric(mort$AgeNew) # Coerce from character to numeric data mode
head(mort$AgeNew, n = 10)
mort$AgeNew <- mort$AgeNew + 1
head(mort$AgeNew, n = 10)
```

## Data Frame Manipulation

Let's create a new column to hold information on whether bighorn age is < 5:

```{r}
below_five <- mort$Age < 5
head(below_five)
```

We've created a simple vector of TRUE/FALSE values. To add it to our data frame, we can use the `cbind()` function, which binds columns together. In this case, it combines our existing data frame with the new logical vector as an additional column.

```{r}
mort <- cbind(mort, below_five)
head(cbind(mort, below_five), 3)
head(mort, 3)
```

If we try to add the below_five vector to our data frame R will return an error. This is because the below_average vector has a different number of elements than rows in the mort data frame. The number of entries in the below_five vector must match the number of rows in the mort data fram to align correctly.

```{r error=FALSE, eval=FALSE}
below_five <- c(TRUE, TRUE, TRUE, TRUE, TRUE)
head(cbind(mort, below_five))
```

Now let's look at how we might add rows to a data frame. 

```{r error=FALSE, eval=FALSE}
# first let's make a mortality dataframe of 3 columns
mort_short <- mort[,c("Animal_ID", "Age", "Cause_Mort")]
new_row <- list('S781', 3, "Lion") 
# note that we have to make the row a list, since each column has a different data type
mort_new <- rbind(mort_short, new_row)
tail(mort_new) # lets you look at the last 6 lines of data
```

***

**`EXERCISE`**

1. Load the bighorn mort table from the bighorn database and assign it the variable name of "mort"
2. Print the names of the mort dataframe using `names()` and `dput(names(mort))`. Why might using dput(names(mort)) be helpful?
2. Select these columns: DeadDate, Animal_ID, COM, Sex, Age, Herd, Recorder
3. Subset to just females using three different methods of subsetting: `subset()`, which indexing, and `filter()`. Ensure that each method provides the same results.
4. Subset to herds that are not Cathedral or Big Arroyo. You will need to use `!` and the `%in%`. Any subsetting method is ok to use.
4. Subset to herds in the Southern Recovery Unit: Mt. Baxter, Mt. Langley, Taboose Creek, Sawmill Canyon, Bubbs Creek, Mt. Williamson. You will need to use `%in%`. Any subsetting method is ok to use
5. Subset to only lion caused mortalities. Use the `unique()` function on the column `COM` to print the category names in the `COM` column. This will provide information on the correct syntax to use for subsetting. For example, is it "Lion", "lion", or "mountain lion". Syntax matters in R. 
6. How many rows and columns are in the mort data frame. 
8. Use `View()` to view you data frame

***
